Historian: accurate reconstruction of ancestral sequences and evolutionary rates

### 2016-11-01 doc/main.tex
Updated

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{gp120tree.pdf}
  \caption{
    Tree of selected HIV and SIV GP120 domains, used for the simulation benchmark.
  }
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{sim/results.pdf}
  \caption{
    Results of the simulation benchmark.
    For each method evaluated, the distribution of the ratio of inferred to true
    indel rate is shown, along with the mean and root-mean-squared error for this
    ratio (which, for a perfect rate inference, would be equal to 1).
    Tool versions are as shown in Table~1.
  }
\end{figure}

@@

From the simulated sequence data, we estimated indel rates
(a) using the true evolutionary alignment (supplied to Historian);
(b) using Historian (in alignment-free mode);
and
(c,d,e) using Prank, ProbCons and Muscle,
the best-performing tools from the ProtPal benchmark \citep{Westesson2012-zg}.
In each case, we estimated insertion and deletion rates $(\hat{\lambda},\hat{\mu})$ and computed relative errors $(\frac{\hat{\lambda}-\lambda}{\lambda},\frac{\hat{\mu}-\mu}{\mu})$
compared to the simulated indel rates.
Having performed this experiment
using the rates estimated for GP120 ($\sigma=1,\lambda=\mu=0.028$),
we repeated the simulation, varying the rate parameters
($\sigma=2,\lambda=\mu=0.056;
\sigma=5,\lambda=\mu=0.14;
\sigma=2,\lambda=\mu=0.112;
\sigma=5,\lambda=\mu=0.28;
\sigma=5,\lambda=\mu=0.005;
\sigma=5,\lambda=\mu=0.01$)
while still using the tree of Figure~1.
Varying rate parameters did not appear
significantly to affect the ranking of the programs by accuracy.

The results of this experiment are summarized in Figure~2.
Note that even perfect knowledge of the true evolutionary alignment does not guarantee perfect reconstruction of the underlying indel rate parameters.
Indel events can overlap and thus be under-counted,
leading to a small negative bias in the estimated rate,
and stochastic noise from a finite sample leads to a spread in the distribution
of estimated rates from individual alignments.
Historian and Prank (which both explicitly aim to provide ancestral reconstructions)
are the most accurate methods for rate estimation,
with a slight edge over ProbCons and Muscle.

As well as varying the rate parameters,
we also repeated the benchmark on a symmetric 8-taxon binary tree.
This simulation yielded similar patterns:
Historian and Prank are still comparable (with a slight edge to Prank in the
symmetric tree, {\em versus} Historian in the GP120 tree),
followed by ProbCons, followed by Muscle.
As with the tree of Figure~1,
varying rate parameters did not appear to change the ranking of the programs by accuracy.

\begin{table}
  \begin{tabular}{r|ccc}
    & Mean SPS & Mean TCS & Notes \\
    \hline
Historian v1.0 & 0.82 & 0.50 & \\
ClustalW & 0.79 & 0.45 & From {\tt drive5.com} \\
Prank v.100701 & 0.78 & 0.44 \\
%MAFFT & 0.87 & 0.61 \\
ProbCons v1.12 & 0.88 & 0.62 & From {\tt drive5.com} \\
Muscle v3.8.31 & 0.89 & 0.64 & 
  \end{tabular}
  \caption{
    Comparison of Historian to other alignment programs using the BAliBase 3.0 benchmark
    with SPS and TCS alignment quality scores \citep{ThompsonEtAl2005}.
%    Average runtime per alignment was 55s for BAliBase and 39s for PREFAB (4 GHz Intel Core i7).
%    Data for ClustalW, Prank (+F option, ClustalW tree), MAFFT (v6.603), ProbCons (v1.12) and Muscle (version 4.0) are from {\tt http://drive5.com/bench/} \citep{Edgar2010}.
  }
\end{table}

Table~2 summarizes an evaluation of Historian
on the BAliBase 3.0 benchmark,
compiled using 3D structural alignments.
%
In general, Historian performs better on these structure-derived benchmarks than Prank,
which also performs ancestral reconstruction \citep{LoytynojaGoldman2008}.
Compared to leading protein aligners that do not attempt ancestral reconstruction,
Historian outperforms ClustalW \citep{LarkinEtAl2007}, but not Muscle \citep{Edgar2004b}
or ProbCons \citep{DoEtAl2005}. %, or MAFFT \citep{KatohEtAl2005}.

\begin{table}
  \begin{tabular}{r|c}
    & Average run time per BAliBase 3.0 alignment \\
    \hline
Historian v1.0 & 55s \\
Prank v.100701 & 520s \\
Muscle v3.8.31 & 1.9s \\
\end{tabular}
\caption{
    Comparison of runtimes of Historian, Prank and Muscle on the BAliBase 3.0 benchmark.
  }
\end{table}

>>>>

The principal parameter determining Historian's execution speed is the number of suboptimal alignments
it retains while building the progressive ancestral reconstruction. Historian provides a command-line option {\tt -fast} which substantially reduces this number of alignments.
We here report results for both default- and fast-mode operation.

@@

\begin{figure}
  \includegraphics[width=.8\columnwidth]{gp120tree.pdf}
  \caption{
    Tree of selected HIV and SIV GP120 domains, used for the simulation benchmark.
  }
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{sim/results.pdf}
  \caption{
    Results of the simulation benchmark.
    For each method evaluated, the distribution of the ratio of inferred to true
    indel rate is shown; for a perfect rate inference, this ratio would be equal to 1.
    The mean of the ratio is also shown; this represents the the systematic error,
    so e.g. using Muscle results in a systematic 15\% underestimate of indels
    in these simulations.
    The root-mean-squared value of the ratio is also shown,
    and the median and interquartile range are annotated above each plot.
    Tool versions are as shown in Table~1.
  }
\end{figure}

@@

From the simulated sequence data, we estimated indel rates
(a) using the true evolutionary alignment (supplied to Historian);
(b,c) using Historian (alignment-free, default and fast modes);
and
(d,e,f) using Prank, ProbCons and Muscle,
the best-performing tools from the ProtPal benchmark \citep{Westesson2012-zg}.
In each case, we estimated insertion and deletion rates $(\hat{\lambda},\hat{\mu})$ and computed relative errors $(\frac{\hat{\lambda}-\lambda}{\lambda},\frac{\hat{\mu}-\mu}{\mu})$
compared to the simulated indel rates.
Having performed this experiment
using the rates estimated for GP120 ($\sigma=1,\lambda=\mu=0.028$),
we repeated the simulation, varying the rate parameters
($\sigma=2,\lambda=\mu=0.056;
\sigma=5,\lambda=\mu=0.14;
\sigma=2,\lambda=\mu=0.112;
\sigma=5,\lambda=\mu=0.28;
\sigma=5,\lambda=\mu=0.005;
\sigma=5,\lambda=\mu=0.01$)
while still using the tree of Figure~1.
Varying rate parameters did not appear
significantly to affect the ranking of the programs.

The results of this experiment are summarized in Figure~2.
Note that even perfect knowledge of the true evolutionary alignment does not guarantee perfect reconstruction of the underlying indel rate parameters.
Indel events can overlap and thus be under-counted,
leading to a small negative bias in the estimated rate,
and stochastic noise from a finite sample leads to a spread in the distribution
of estimated rates from individual alignments.
Historian and Prank (which both explicitly aim to provide ancestral reconstructions)
are the most accurate methods for rate estimation,
with a slight edge over ProbCons and Muscle.
Historian's default mode has a slight edge over its fast mode in rate estimation accuracy.

As well as varying the rate parameters,
we also repeated the benchmark on a symmetric 8-taxon binary tree.
This simulation yielded similar patterns:
Historian and Prank are still comparable (with a slight edge to Prank in the
symmetric tree, {\em versus} Historian in the GP120 tree),
followed by ProbCons, followed by Muscle.
As with the tree of Figure~1,
varying simulation parameters did not appear to change the ranking.

\begin{table}
  \begin{tabular}{r|ccc}
    & Mean SPS & Mean TCS & Notes \\
    \hline
    Historian v1.0.2 & 0.822 & 0.497 & \\
    (fast mode) & 0.820 & 0.494 & \\
ClustalW & 0.787 & 0.447 & From {\tt drive5.com} \\
Prank v.100701 & 0.784 & 0.441 \\
%MAFFT & 0.873 & 0.607 \\
ProbCons v1.12 & 0.883 & 0.619 & From {\tt drive5.com} \\
Muscle v3.8.31 & 0.843 & 0.532 & \\
Muscle v4.0 & 0.889 & 0.642 &  From {\tt drive5.com}
  \end{tabular}
  \caption{
    Comparison of Historian to other alignment programs using the BAliBase 3.0 benchmark
    with SPS and TCS alignment quality scores \citep{ThompsonEtAl2005}.
%    Average runtime per alignment was 55s for BAliBase and 39s for PREFAB (4 GHz Intel Core i7).
%    Data for ClustalW, Prank (+F option, ClustalW tree), MAFFT (v6.603), ProbCons (v1.12) and Muscle (version 4.0) are from {\tt http://drive5.com/bench/} \citep{Edgar2010}.
  }
\end{table}

Table~2 summarizes an evaluation of Historian
on the BAliBase 3.0 benchmark,
compiled using 3D structural alignments.
%
In general, Historian performs better on these structure-derived benchmarks than Prank,
which also performs ancestral reconstruction \citep{LoytynojaGoldman2008}.
Compared to leading protein aligners that do not attempt ancestral reconstruction,
Historian outperforms ClustalW \citep{LarkinEtAl2007}, but not Muscle \citep{Edgar2004b}
or ProbCons \citep{DoEtAl2005}. %, or MAFFT \citep{KatohEtAl2005}.
No significant difference was found between Historian's default and fast modes.

\begin{table}
  \begin{tabular}{r|c}
    & Average run time per alignment \\
    \hline
Historian v1.0.2 & 233s \\
 (fast mode) & 55s \\
Prank v.100701 & 520s \\
Muscle v3.8.31 & 1.9s \\
\end{tabular}
\caption{
    Comparison of runtimes of Historian, Prank and Muscle on the BAliBase 3.0 benchmark.
  }
\end{table}
